### 2019-06-05 chapter_00_open_science.tex
Minor corrections, exec black and isort on python scripts

The Free and Open-Source Software (FOSS) movement has given rise to the tools
that enable open-source software development today.
%
The term \emph{free} in FOSS is a misnomer, as it actually stands for
\emph{freedom} (to use, modify and distribute). The FOSS movement has
dramatically decreased the cost of using computers, which is evident from the
widespread use of GNU/Linux systems in desktops, computing clusters and
web-servers in academia and beyond.
%
The FOSS culture can be traced back to the success of the type-setting
standards \TeX\ (1977) and \LaTeX\ (1985) for authoring scientific publications.
%
Another founding moment for the FOSS movement was the launch of the GNU project
in 1983 by \href{https://en.wikipedia.org/wiki/Richard_Stallman}{Richard
  Stallman}, to create a Unix-like computer operating system composed entirely
  of free software\footnote{For completeness, see also the work done on
  \href{https://www.levenez.com/unix/}{other Unix operating systems, for
example BSD}.}.
%
GNU is today known for its compiler collection (GCC) and a multitude of tools
which combined with the Linux kernel (created by
\href{https://en.wikipedia.org/wiki/Linus_Torvalds}{Linus Torvalds} in 1991)
forms the GNU/Linux operating system
% and its several distributions (Debian, Red Hat, Slackware, Arch Linux and
% Gentoo to name a few)
that we are familiar with today.
%
% FOSS movement has had huge successes in many frontiers --- Apache could be
% termed as the ``first killer-app of Linux'' and now with an ever-increasing
% suite of softwares including Firefox, LibreOffice, Gimp as solid alternatives
% to proprietary offerings.
%
Linux has become the most widely used kernel, being deployed on servers,
personal computers, embedded devices, and also smart phones (with Android). The
success of the Linux kernel project was attributed to the \emph{bazaar} model of
software development, as described in \citet{raymond_cathedral_1999}, wherein a
code can stay structured despite releasing frequently and delegating tasks to
community with myriad agendas.

@@

The Python community has thrived remarkably in the last few years, which is
evident from the fact that by the end of 2017, Python became the second most
popular language in terms of pull-requests in GitHub\footnote{%
%See GitHub's annual report titled \href
\url{https://octoverse.github.com/}
% { ``The State of the Octoverse 2017''}.
%
} and
\fnref%
% [See the blog post ``The Incredible Growth of Python'']%
{https://stackoverflow.blog/2017/09/06/incredible-growth-python/}%
{questions on StackOverflow}.
%
TIOBE-index\footnote{\url{https://www.tiobe.com/tiobe-index}}, which assess the
popularity of a language based on number of skilled users, courses and
third-party vendors, ranks Python at number four. IEEE Spectrum\footnote{
%See the analysis titled 
\url{https://spectrum.ieee.org/computing/software/the-2017-top-programming-languages}
%{``The 2017 Top Programming Languages''} on IEEE spectrum.
}
ranks Python as the most popular language in 2017 with a metric
based on trends, demand by employers and users on open-source hubs.  All the
rankings agree that Python usage has been steadily growing and is here to stay.
Python largely outclasses languages like Fortran and Matlab according to these
statistics, and is now clearly one of the mainstream languages with C, Java,
C++ and JavaScript.
%
The language has been designed to boost the communication of technical ideas
between humans. 
%
Let us summarize some characteristics of the Python languages that have lead to
such an incredible success.
%
Python differs from other mainstream languages for a variety of reasons:

@@

Python has a very large and supportive community (see, for example,
\fnref{http://stackoverflow.com/tags/python}{Stack Overflow tags}).
%
Python development is also supported by companies, and most of them specialize
in web technologies and data-driven research. However, recently, companies
involved in scientific software development have started to become quite
influential. For example, Anaconda Inc.\ is strongly engaged in the development
of tools like JIT Python accelerator \pack{Numba}, and parallelism framework
\pack{dask}. Hardware giants such as Intel and Nvidia also contribute to the
Python scientific ecosystem.
%
To summarize, Python's many features and its adoption in several domains
makes it a sensible language choice for scientific purposes.

@@

\item \textbf{Concurrent, but not parallel multi-threading.} Threads behave
	like light sub-programs. In principle, they can use the different cores
	of the CPU at the same time, but the CPython interpreter forbid threads
	to interpret python code at the same time\footnote{See
	\url{https://opensource.com/article/17/4/grok-gil}
	%{a blog article ``How to grok the GIL''}
	% and \url{https://faster-cpython.readthedocs.io/cpython37.html}{a
	% technical presentation by a CPython core developer}.
	}. In other words, threads can coexist concurrently, by not execute in
	parallel.
%
This limitation comes from an important detail in the implementation of the
CPython interpreter: a Global Interpreter Lock (GIL) to prevent race conditions
that could corrupt data.
%
GIL greatly simplifies the implementation of CPython and that
it is very difficult to remove it while keeping other nice technical properties
of CPython intact\footnote{See
\url{https://wiki.python.org/moin/GlobalInterpreterLock}. Note that PyPy
also uses a GIL, while two other implementations of the Python language, Jython
and IronPython, do not have this limitation}.

>>>>

The Free and Open-Source Software (FOSS) movement has given rise to the tools
that enable open-source software development today.
%
The term \emph{free} in FOSS is a misnomer, as it actually stands for
\emph{freedom} (to use, modify and distribute). The FOSS movement has
dramatically decreased the cost of using computers, which is evident from the
widespread use of GNU/Linux systems in desktops, computing clusters and
web-servers in academia and beyond.
%
The FOSS culture can be traced back to the success of typesetting
standards \TeX\ (1977) and \LaTeX\ (1985) for authoring scientific publications.
%
Another founding moment for the FOSS movement was the launch of the GNU project
in 1983 by \href{https://en.wikipedia.org/wiki/Richard_Stallman}{Richard
  Stallman}, to create a Unix-like computer operating system composed entirely
  of free software\footnote{For completeness, see also the work done on
  \href{https://www.levenez.com/unix/}{other Unix operating systems, for
example BSD}.}.
%
GNU is today known for its compiler collection (GCC) and a multitude of tools
which combined with the Linux kernel (created by
\href{https://en.wikipedia.org/wiki/Linus_Torvalds}{Linus Torvalds} in 1991)
forms the GNU/Linux operating system
% and its several distributions (Debian, Red Hat, Slackware, Arch Linux and
% Gentoo to name a few)
that we are familiar with today.
%
% FOSS movement has had huge successes in many frontiers --- Apache could be
% termed as the ``first killer-app of Linux'' and now with an ever-increasing
% suite of softwares including Firefox, LibreOffice, Gimp as solid alternatives
% to proprietary offerings.
%
Linux has become the most widely used kernel, being deployed on servers,
personal computers, embedded devices, and also smart phones (with Android). The
success of the Linux kernel project was attributed to the \emph{bazaar} model of
software development, as described in \citet{raymond_cathedral_1999}, wherein a
code can stay structured despite releasing frequently and delegating tasks to
community with myriad agendas.

@@

The Python community has thrived remarkably in the last few years, which is
evident from the fact that by the end of 2017, Python became the second most
popular language in terms of pull-requests in GitHub\footnote{%
%See GitHub's annual report titled \href
\url{https://octoverse.github.com/2017}
% { ``The State of the Octoverse 2017''}.
%
} and
\fnref%
% [See the blog post ``The Incredible Growth of Python'']%
{https://stackoverflow.blog/2017/09/06/incredible-growth-python/}%
{questions on StackOverflow}.
%
TIOBE-index\footnote{\url{https://www.tiobe.com/tiobe-index}}, which assess the
popularity of a language based on number of skilled users, courses and
third-party vendors, ranks Python at number four. IEEE Spectrum\footnote{
%See the analysis titled 
\url{https://spectrum.ieee.org/computing/software/the-2017-top-programming-languages}
%{``The 2017 Top Programming Languages''} on IEEE spectrum.
}
ranks Python as the most popular language in 2017 with a metric
based on trends, demand by employers and users on open-source hubs.  All the
rankings agree that Python usage has been steadily growing and is here to stay.
Python largely outclasses languages like Fortran and Matlab according to these
statistics, and is now clearly one of the mainstream languages with C, Java,
C++ and JavaScript.
%
The language has been designed to boost the communication of technical ideas
between humans. 
%
Let us summarize some characteristics of the Python languages that have lead to
such an incredible success.
%
Python differs from other mainstream languages for a variety of reasons:

@@

Python has a very large and supportive community (see, for example,
\fnref{http://stackoverflow.com/tags/python}{Stack Overflow tags}).
%
Python development is also supported by companies, and most of them specialize
in web technologies and data-driven research. However, recently, companies
involved in scientific software development have started to become quite
influential. For example, Anaconda Inc.\ is strongly engaged in the development
of tools like JIT Python accelerator \pack{Numba}, and parallelism framework
\pack{dask}. Hardware giants such as Intel and Nvidia also contribute to the
Python scientific ecosystem.
%
To summarize, Python's numerous features and its adoption in several domains
makes it a sensible language choice for scientific purposes.

@@

\item \textbf{Concurrent, but not parallel multi-threading.} Threads behave
	like light sub-programs. In principle, they can use the different cores
	of the CPU at the same time, but the CPython interpreter forbid threads
	to interpret python code at the same time\footnote{See
	\url{https://opensource.com/article/17/4/grok-gil}
	%{a blog article ``How to grok the GIL''}
	% and \url{https://faster-cpython.readthedocs.io/cpython37.html}{a
	% technical presentation by a CPython core developer}.
	}. In other words, threads can coexist concurrently, by not executing in
	parallel.
%
This limitation comes from an important detail in the implementation of the
CPython interpreter: a Global Interpreter Lock (GIL) to prevent race conditions
that could corrupt data.
%
GIL greatly simplifies the implementation of CPython and that
it is very difficult to remove it while keeping other nice technical properties
of CPython intact\footnote{See
\url{https://wiki.python.org/moin/GlobalInterpreterLock}. Note that PyPy
also uses a GIL, while two other implementations of the Python language, Jython
and IronPython, do not have this limitation}.
